---
title: "US Voting Patterns Analysis"
author: "Emma Bonanno, Emily Johansen, Katia Krotova"
date: "10/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Research Question

Are there significant correlations between rural/suburban/urban characteristics, population density, race, income, and county-level voting patterns in the 2020 presidential election? 


## Prior Research

Gimpel et. al (2020) find that urban–rural political party splits persist even after accounting for an array of individual-level characteristics, such as race, income and education. Hill et. al (2019) demonstrate that at the precinct level, local demographic changes and increasing diversity did not correlate to an increase in support for anti-immigration Republican candidates in the lead up to the 2016 presidential election. 


```{r, load packagaes and ACS, include = FALSE, quiet = TRUE}
library(tidyverse)
library(tidycensus)
library(readxl)
library(knitr)
library(tigris)
library(sf)
library(gridExtra)
library(jtools)
vars_2017 <- load_variables(2017, "acs5")
```
## Data

Our sample population for this analysis is the full set of all counties across the United States (3,220 observations total). 

The data included in the analysis is:


* People per square mile (American Community Survey 2017)
* Median income (American Community Survey 2017)
* Percentage of population identifying as White only and Black / African-American only (American Community Survey 2017)
* Percentage of population voting Republican in the 2020 presidential election (MIT Election Data and Science Lab, 2018)
* Urban-Rural County classification (National Center for Health Statistics 2013)

Load ACS Data
```{r, quiet = TRUE, results= FALSE}
acs <- get_acs(geography = "county", 
                        year = 2017,
                        variables = c(total_pop = "B01003_001",
                                      med_inc = "B06011_001",
                                      Black = "B02001_003",
                                      white = "B02001_002"),
                        output = "wide",
                        geometry = TRUE)
areas <- counties() %>%
  st_set_geometry(NULL) %>%
  mutate(sq_miles = ALAND / 2589988.11) %>%
  select(GEOID, sq_miles)
```
Load suburban-rural classification
```{r}
CO_type <- read_xlsx(path = "data/NCHSURCodes2013.xlsx", 
                      sheet = "NCHSURCodes2013") %>%
  mutate(GEOID = case_when(str_length(as.character(`FIPS code`)) == 5 ~ 
                            as.character(`FIPS code`),
                          str_length(as.character(`FIPS code`)) == 4 ~
                            paste("0", `FIPS code`, sep=""),
                          TRUE ~ "unknown")) %>%
  mutate(type = case_when(`2013 code` == 1 ~ "Large central metro",
                          `2013 code` == 2 ~ "Large fringe metro",
                          `2013 code` == 3 ~ "Medium metro",
                          `2013 code` == 4 ~ "Small metro",
                          `2013 code` == 5 ~ "Micropolitan",
                          `2013 code` == 6 ~ "Non-core",
                          TRUE ~ "unknown")) %>%
  select(GEOID, type)
```

Load election result (Democrat/Republican) data 
```{r, quiet = TRUE, results = 'hide'}
election <- read_csv('data/countypres_2000-2020.csv') %>%
  filter(year == 2020) %>%
  filter(party == "REPUBLICAN") %>%
  rename(GEOID = county_fips) %>%
  group_by(GEOID) %>%
  summarize(candidatevotes = sum(candidatevotes),
            totalvotes = first(totalvotes)) %>%
  mutate(pct_GOP = (candidatevotes / totalvotes)*100) %>%
  select(GEOID, pct_GOP)
```

```{r}
election <- election %>%
   mutate(GEOID = as.character(GEOID))
```
Join datasets: 
```{r}
data <- left_join(acs, areas) %>%
  left_join(election) %>%
  left_join(CO_type) %>%
  mutate(pop_dens = total_popE / sq_miles) %>%
  mutate(black_pct = (BlackE / total_popE)*100) %>%
  mutate(white_pct = (whiteE / total_popE)*100) %>%
  select(GEOID, NAME, total_popE, pop_dens, med_incE, BlackE, black_pct, whiteE, white_pct, type, pct_GOP)

kable(head(data))
```
# Descriptive Statistics

## Continuous Variables

The continuous variable used in this research are population density, median income, percentage of population identifying as White-only, percentage of population identifying as Black or African American-only, and percentage of a county's population that voted Republican in the 2020 presidential election. 

For each continuous variable, calculate:

1. The sample mean
2. The 95-percent confidence interval for the population mean
3. The sample standard deviation
4. The interquartile range
5. An illustration (e.g. a histogram) and description of the distribution.

### Calculations

```{r, the sample mean and the 95-percent confidence interval from the population mean by using one sample t-test}
medinc_t_test <- t.test(data$med_incE)
dens_t_test <- t.test(data$pop_dens)
black_t_test <- t.test(data$black_pct)
white_t_test <- t.test(data$white_pct)
GOP_t_test <- t.test(data$pct_GOP)
```

```{r, calculationg the median and the interquartile range using the quantile function}
medinc_quartiles <- quantile(data$med_incE, na.rm = TRUE)
dens_quartiles <- quantile(data$pop_dens, na.rm = TRUE)
black_quartiles <- quantile(data$black_pct, na.rm = TRUE)
white_quartiles <- quantile(data$white_pct, na.rm = TRUE)
GOP_quartiles <- quantile(data$pct_GOP, na.rm = TRUE)

```

```{r, calculationg standard deviation}
medinc_st_dev <- sd(data$med_incE, na.rm = TRUE)
dens_st_dev <- sd(data$pop_dens, na.rm = TRUE)
black_st_dev <- sd(data$black_pct, na.rm = TRUE)
white_st_dev <- sd(data$white_pct, na.rm = TRUE)
GOP_st_dev <- sd(data$pct_GOP, na.rm = TRUE)

```

```{r, generationg histogram for each variable}
medinc_hist <- ggplot(data) +
  geom_histogram(aes(x = med_incE),
                 bins = 30)
dens_hist <- ggplot(data) +
  geom_histogram(aes(x = pop_dens),
                 bins = 30) +
  scale_x_continuous(trans = "log")
black_hist <- ggplot(data) +
  geom_histogram(aes(x = black_pct),
                 bins = 30) +
   scale_x_continuous(trans = "log")
white_hist <- ggplot(data) +
  geom_histogram(aes(x = white_pct),
                 bins = 30)+
  scale_x_continuous(trans = "log")

GOP_hist <- ggplot(data) +
  geom_histogram(aes(x = pct_GOP),
                 bins = 30) +
  scale_x_continuous(trans = "log")

```


### Displaying Results

```{r, displaying results from the continous variables, warning=FALSE}

medinc_t_test
medinc_quartiles
medinc_st_dev
medinc_hist
```
```{r}
cont_summary <- tibble(
  Variable = c("Median income", 
               "Population density (people per square mile)",
               "Percentage of population identifying as Black or African-American only",
               "Percentage of population identifying as White only",
               "Percent who voted Republican"),
  `Sample mean` = c(medinc_t_test$estimate,
                    dens_t_test$estimate,
                    white_t_test$estimate,
                    black_t_test$estimate,
                    GOP_t_test$estimate),
  `Population mean (95% confidence) - low` = 
    c(medinc_t_test$conf.int[1],
      dens_t_test$conf.int[1],
       white_t_test$conf.int[1],
      black_t_test$conf.int[1],
      GOP_t_test$conf.int[1]),
  `Population mean (95% confidence) - high` =
    c(medinc_t_test$conf.int[2],
      dens_t_test$conf.int[2],
      white_t_test$conf.int[2],
      black_t_test$conf.int[2],
      GOP_t_test$conf.int[2]),
  Median = c(medinc_quartiles[3],
             dens_quartiles[3],
             white_quartiles[3],
             black_quartiles[3],
             GOP_quartiles[3]),
  `Interquartile range` = c(medinc_quartiles[4] - medinc_quartiles[2],
                            dens_quartiles[4] - dens_quartiles[2],
                            white_quartiles[4] - white_quartiles[2],
                            black_quartiles[4] - black_quartiles[2],
                            GOP_quartiles[4] - GOP_quartiles[2]),
  `Standard deviation` = c(medinc_st_dev,
                          dens_st_dev,
                          white_st_dev,
                          black_st_dev,
                          GOP_st_dev))

kable(cont_summary, digits = 0)
```
```{r}
pretty_medinc_hist <- medinc_hist +
  theme_bw() +
  scale_x_continuous(name = "Median income") +
  scale_y_continuous(name = "Number of counties") +
  theme(axis.text.x = element_text(angle = 90))


pretty_GOP_hist <- GOP_hist +
  theme_bw() +
  scale_x_continuous(name = "% of county residents who \n voted Republican") +
  scale_y_continuous(name = "Number of counties") +
  theme(axis.text.x = element_text(angle = 90))

pretty_dens_hist <- dens_hist +
  theme_bw() + 
  scale_x_continuous(name = "Population density\n(residents per square mile)",
                     trans = "log",
                     breaks = c(0.1, 1, 10, 100, 1000, 10000),
                     labels = c("0.1", "1", "10", "100", "1000", "10,000")) +
  scale_y_continuous(name = "Number of counties") +
  theme(axis.text.x = element_text(angle = 90))

pretty_white_hist = white_hist +
  theme_bw() +
  scale_x_continuous(name = "Percentage of population\nidentifying as White only") + 
  scale_y_continuous(name = "Number of counties") +
  theme(axis.text.x = element_text(angle = 90)) 

pretty_black_hist = black_hist +
  theme_bw() +
  scale_x_continuous(name = "Percentage of population\nidentifying as Black or\nAfrican-American only") +
  scale_y_continuous(name = "Number of counties") +
  theme(axis.text.x = element_text(angle = 90))

grid.arrange(pretty_medinc_hist, pretty_dens_hist, pretty_white_hist, pretty_black_hist, pretty_GOP_hist,
             ncol = 3)
```

## Categorical Variables

The categorical variable used in this analysis is:

* The county type (non-core, micropolitan, metropolitan, etc), based on the NCHS classification.

### Calculations

```{r}

pct_large_metro <- t.test(data$type == "Large central metro")
pct_fringe_metro <- t.test(data$type == "Large fringe metro")
pct_med_metro <- t.test(data$type == "Medium metro")
pct_small_metro <- t.test(data$type == "Small metro")
pct_micro <- t.test(data$type == "Micropolitan")
pct_rural <- t.test(data$type == "Non-core")

```

```{r}
cat_summary_type <- tibble(`County type` = 
                              c("Large central metro",
                                "Large fringe metro",
                                "Medium metro",
                                "Small metro",
                                "Micropolitan",
                                "Non-core"),
                            `Sample proportion` = 
                              c(pct_large_metro$estimate * 100,
                                pct_fringe_metro$estimate * 100,
                                pct_med_metro$estimate * 100,
                                pct_small_metro$estimate * 100,
                                pct_micro$estimate * 100,
                                pct_rural$estimate * 100),
                            `95-percent confidence - low` = 
                              c(pct_large_metro$conf.int[1] * 100,
                                pct_fringe_metro$conf.int[1] * 100,
                                pct_med_metro$conf.int[1] * 100,
                                pct_small_metro$conf.int[1] * 100,
                                pct_micro$conf.int[1] * 100,
                                pct_rural$conf.int[1] * 100),
                            `95-percent confidence - high` = 
                              c(pct_large_metro$conf.int[2] * 100,
                                pct_fringe_metro$conf.int[2] * 100,
                                pct_med_metro$conf.int[2] * 100,
                                pct_small_metro$conf.int[2] * 100,
                                pct_micro$conf.int[2] * 100,
                                pct_rural$conf.int[2] * 100))
kable(cat_summary_type, digits = 0)
```

```{r}
ggplot(cat_summary_type) +
  geom_bar(aes(x = `County type`, 
               y = `Sample proportion`),
           stat = "identity") +
  geom_errorbar(aes(x = `County type`, 
               y = `Sample proportion`,
               ymin = `95-percent confidence - low`,
               ymax = `95-percent confidence - high`),
           stat = "identity") +
  scale_y_continuous(name = "Percent of counties",
                     breaks = c(0, 10, 20, 30, 40),
                     labels = c("0", "10%", "20%", "30%", "40%")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90))
```

# Bivariate Regression on Continuous Variables

## Population Density

```{r}
cor.test(~ pct_GOP + pop_dens, data = data)
```

```{r}
dens_model <- lm(pct_GOP ~ pop_dens, data = data)

summary(dens_model)
```
The 95-percent confidence interval for the correlation does not include zero, and the correlation coefficient is negative. This means we can say with 95-percent confidence that higher population density within a county is associated with a lower percentage of Republican voters.

##  Median Income 

```{r}
cor.test(~ pct_GOP + med_incE, data = data)

inc_model <- lm(pct_GOP ~ med_incE, data = data)

summary(inc_model)
```
The 95-percent confidence interval for the correlation does not include zero, and the correlation coefficient is negative. This means we can say with 95-percent confidence that higher median income within a county is associated with a lower percentage of Republican voters.


## Percentage of population identifying as white only

```{r}
cor.test(~ pct_GOP + white_pct, data = data)

inc_model <- lm(pct_GOP ~ white_pct, data = data)

summary(inc_model)
```
The 95-percent confidence interval for the correlation does not include zero, and the correlation coefficient is positive. This means we can say with 95-percent confidence that a higher percentage of white-only identifying people within a county is associated with a higher percentage of Republican voters.


## Percent of population identifying as Black / African-American only 

```{r}
cor.test(~ pct_GOP + black_pct, data = data)

inc_model <- lm(pct_GOP ~ black_pct, data = data)

summary(inc_model)
```
The 95-percent confidence interval for the correlation does not include zero, and the correlation coefficient is negative. This means we can say with 95-percent confidence that a higher percentage of Black-only identifying people within a county is associated with a lower percentage of Republican voters.


# Bivariate Regression on Categorical Variables

## Urban-Rural County classification

Is there a relationship between voting patterns and the urban/rural makeup of a county?

```{r}
data <- data %>%
  mutate(type = as.factor(type)) %>%
  mutate(type = relevel(type, "Large central metro"))
type_model <- lm(pct_GOP ~ type, data)
summary(type_model)
```
Since the coefficient for each is positive (represented in the first "Estimate Std" column), there is a positive relationship between each of these types and GOP voting. Since the coefficients are higher for micropolitan and non-core cities, we can infer that GOP voting is generally higher in these types of cities than in the others which have lower coefficients. 

# Multivariate Regression
```{r}

full_model <- lm(pct_GOP ~ med_incE + pop_dens + white_pct + black_pct + type, data)

summary(full_model)

```
## Results

Since all of the variables have a p-value of less than 0.05, they are all significant at a 95% confidence interval. 

**Median Income:** We found that an increase in income corresponded with a decrease in the percentage of voters in a county who voted for GOP in the 2020 election. In the bivariate regression a one dollar increase was associated with a decrease in the percent of GOP voters of 0.00056%. If we control for all other variables, a one dollar increase in income was relatively similar, associated with a decrease in the GOP percent of 0.00059%. 

**Population Density:** A one person/sq. mi. increase in population density corresponded with a -0.0021766% change in GOP voters. In the multivariate regression, once we controlled for other factors, a one person increase in population density was associated with a decrease of 0.00055% in GOP voters. As a result, our analysis suggested that population density had a much weaker effect on the percentage of voters in a county who voted for the GOP when all other variables were held constant. 

**White Pct:** Based on our analysis, whiteness alone was not a good predictor of a county's voting patterns. In our bivariate regression, we found that a 1% increase in the white population of a county was associated with an increase in the GOP % of 0.53130. On the other hand, when all other variables were held constant, we found that a similar increase in the white population corresponding with an increase in votes won by the GOP of 0.4555%.

**Black Pct:** In the bivariate regression a one percent increase in the county's Black population was associated with a decrease in the percent of GOP voters of 0.50714%. If we control for all other variables, our multivariate regression showed that a one percent increase in Black population was associated with a decrease in the GOP percent of 0.068688%. 

**County Type:** Since most of the county types had similar coefficients, they tend to have a similar effect on GOP voting outcomes. Since population density may account for some of the differences between county types, it may be a better predictor. Still, non-core (rural) and micropolitan counties saw the highest increase in the share of GOP votes.

Overall, our model explains about 45% of the variation in county-level voting patterns.


# Transformations & Scaling

## Mean Centering 
We chose to mean center the population density and median income variables. By doing so, we are able to predict how variation in density and income away from average impacts the percent of GOP votes. 

```{r}

centered_data <- data %>%
  mutate(pop_dens = pop_dens - mean(pop_dens, na.rm=TRUE),
         med_incE = med_incE - mean(med_incE, na.rm=TRUE))

centered_model <- lm(pct_GOP ~ med_incE + pop_dens + white_pct + black_pct + type, centered_data)

export_summs(full_model, centered_model, 
             error_format = "(p = {p.value})",
             error_pos = "same",
             model.names = c("Initial", "Centered"),
             coeffs = "coeff_names")

```
By mean centering the data, the R-squared value remained constant with our initial model, but the intercept shifted. In this case, the mean-centered intercept represents the percentage of GOP votes in a hypothetical Large Central Metro county where population density and median income are average and there are zero white or Black residents.

```{r}
coeff_names <- c("Constant" = "(Intercept)",
                 "Median income" = "med_incE",
                 "Population density (people per square mile)" = "pop_dens",
                 "White Population (%)" = "white_pct",
                 "Black Population (%)" = "black_pct",
                 "Large fringe metro" = "typeLarge fringe metro",
                 "Medium metro" = "typeMedium metro",
                 "Small metro" = "typeSmall metro",
                 "Micropolitan" = "typeMicropolitan",
                 "Rural" = "typeNon-core")

export_summs(full_model, centered_model, 
             error_format = "(p = {p.value})",
             error_pos = "same",
             model.names = c("Initial", "Centered"),
             coefs = coeff_names)
```
## Scaling

We chose to scale the population density and median income variables since the coefficients were extremely small in our initial model. Our initial model predicted the change in the percent GOP votes for a one dollar change in median income and a one person per square mile change in population density. Since the variables being rescaled have already been mean-centered, the new model will predict the change in GOP votes for a $1,000 change in median income over the average, and the change in GOP votes for a 100 person change in population density over the average. 

```{r}
centered_data_scaled <- centered_data %>%
  mutate(med_incE = med_incE / 1000,
         pop_dens_100 = pop_dens / 100)

coeff_names_scaled <- c("Constant" = "(Intercept)",
                 "Median Income ($000s)" = "med_incE",
                 "Population density (100 people per square mile)" = "pop_dens_100",
                 "White Population" = "white_pct",
                 "Black Population" = "black_pct",
                 "Large fringe metro" = "typeLarge fringe metro",
                 "Medium metro" = "typeMedium metro",
                 "Small metro" = "typeSmall metro",
                 "Micropolitan" = "typeMicropolitan",
                 "Rural" = "typeNon-core")
                   
                   
centered_model_scaled <- lm(pct_GOP ~ med_incE + pop_dens_100 + white_pct + black_pct + type, centered_data_scaled)

export_summs(full_model, centered_model,centered_model_scaled,
             error_format = "(p = {p.value})",
             error_pos = "same",
             model.names = c("Initial", 
                             "Centered", 
                             "Centered\n(per 100 residents)"),
             coefs = coeff_names_scaled)
```
By mean centering and scaling our model, we are able to produce more easily interpretable coefficients. Our centered and scaled model now predicts the effect on GOP votes as median income deviates from the average by 1,000 dollars, as well as for a change in population density of 100 people over the average. 

In the initial and centered models, the coefficient for median income was extremely small (displayed as zero) because we were analyzing the change in GOP voting percentage when median income changes by only one dollar. In scaling the data, we were able to identify the change in GOP voting % when median income changes by 1,000. The output of the model was more meaningful in this case and returned a 0.59% decrease in GOP votes for a $1,000 increase in median income over the average. 

Similarly, the coefficient for population density was also extremely small because we were analyzing the change in GOP votes when population density changes by only one person. In scaling the data, we were able to identify the change in GOP voting % when population density changes by 100 people per square mile. The model in this case returned a 0.05% decrease in percent GOP votes when population density increased by 100 over the average. 

Since the R-squared value remained the same for all three models (equal to 0.46), our model fit did not change. This implies that 46% of change in GOP votes (%) can be attributed to these variables. 

# References

Gimpel, J.G., N. Lovin, B. Moy, & A. Reeves (2020). "The Urban–Rural Gulf in American Political Behavior". Political Behavior, 42, 1343-1368.

Hill, Seth J.,  Daniel J. Hopkins, Gregory A. Huber (2019). "Local demographic changes and US presidential voting, 2012 to 2016". Proceedings of the National Academy of Arts and Sciences, December 10, 2019 116 (50) 25023-25028. 